Disparities in Climate-Induced Health Outcomes in the Greater Toronto Area.

### 2022-07-12 outputs/paper/paper.qmd
Updates to paper

pollution %>%
  select(city, pollutant, mean_pollution, day, month, year) %>%
  slice(c(100, 1000, 1200, 1500, 1800, 2000, 2500, 2800, 3600, 5000, 6000, 7000, 10000, 11000, 15000)) %>%
  kable(
    caption = "Extracting rows from the Pollution data from 2002 - 2020",
    col.names=c("City", "Pollutant", "Mean Pollution", "Day", "Month", "Year"),
    linesep = "", digits = 2, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10)

@@

temp_2002_2022 %>% filter(month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar") %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point() + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature")

@@

temp_2002_2022 %>% filter(month_name == "Apr" | month_name == "May" | month_name == "Jun" | month_name == "Jul") %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point() + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature")

@@

temp_2002_2022 %>% filter(month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov") %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point() + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature")

@@

pollution %>% ggplot(
  aes(x = year, y = mean_pollution, color = pollutant)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution, color = pollutant)) + geom_point() + 
  labs(color="Pollutant") + xlab("Month") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = year, y = mean_pollution, factor(month_name, level = level_order))) +
  geom_point() + labs(color="Month") + xlab("Year") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "NO2" ) %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "NO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  xlab("Year") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)")
```

In figure 12 and 13 we see a similar pattern as NO and NO2 where the level of pollution is decreasing throughout the years from 2002 to 2020 and the highest levels are in the winter months. The highest level of pollution for CO was seen in 2002 with 1762.5 ppb. This level is much higher than any of the other pollutants. This suggests that carbon monoxide levels are much more widespread than any of the other pollutants we have analyzed and due to this can have a affect on our health. Even with the decreasing levels, the affects will still be dangerous. The highest level in 2020 was around 550 ppb which is almost 5 times higher than the levels of NO, NO2, and O3 in 2020.

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  xlab("Year") + ylab("Mean Pollution (ppb)")
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)")
```

@@

#### Visualizing Mean Pollution by Temperature, Pollutant and Month

@@

pollu_temp %>% filter((pollutant != "CO") & (month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar")) %>% ggplot(aes(max_temp, mean_pollution, color=factor(month_name, level=level_order))) + geom_point() + facet_grid(. ~ pollutant) + labs(color="Month") + xlab("Maximum Temperature") +  ylab("Mean Pollution")
```

@@

pollu_temp %>% filter((pollutant == "CO") & (month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar")) %>% ggplot(aes(max_temp, mean_pollution, color=factor(month_name, level=level_order))) + geom_point() + facet_grid(. ~ pollutant) + labs(color="Month") + xlab("Maximum Temperature") +  ylab("Mean Pollution")
```

@@

pollu_temp %>% filter((pollutant != "CO") & (month_name == "Apr" | month_name == "May" | month_name == "Jun" |  month_name == "Jul")) %>% ggplot(aes(max_temp, mean_pollution, color=factor(month_name, level=level_order))) + geom_point() + facet_grid(. ~ pollutant) + labs(color="Month") + xlab("Maximum Temperature") + ylab("Mean Pollution")

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for CO by Month, Maximum Temperature, and Pollutant for April to July", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant == "CO") & (month_name == "Apr" | month_name == "May" | month_name == "Jun" |  month_name == "Jul")) %>% ggplot(aes(max_temp, mean_pollution, color=factor(month_name, level=level_order))) + geom_point() + facet_grid(. ~ pollutant) + labs(color="Month") + xlab("Maximum Temperature") + ylab("Mean Pollution")

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for NO, NO2, O3 and SO2 by Month, Maximum Temperature, and Pollutant for August to November", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant != "CO") & (month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov")) %>% ggplot(aes(max_temp, mean_pollution, color=factor(month_name, level=level_order))) + geom_point() + facet_grid(. ~ pollutant) + labs(color="Month") + xlab("Maximum Temperature") + ylab("Mean Pollution")

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for CO by Month, Maximum Temperature, and Pollutant for August to November", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant == "CO") & (month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov")) %>% ggplot(aes(max_temp, mean_pollution, color=factor(month_name, level=level_order))) + geom_point() + facet_grid(. ~ pollutant) + labs(color="Month") + xlab("Maximum Temperature") + ylab("Mean Pollution")

@@

>>>>

pollution %>%
  select(city, pollutant, mean_pollution, day, month, year) %>%
  slice(c(100, 1000, 1200, 1500, 1800, 2000, 2500, 2800, 3600, 5000, 6000, 7000, 10000, 11000, 15000)) %>%
  kable(
    caption = "Extracting rows from the Pollution data from 2002 - 2020",
    col.names=c("City", "Pollutant", "Mean Pollution", "Day", "Month", "Year"),
    linesep = "", digits = 2, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10) + theme_bw()

@@

temp_2002_2022 %>% filter(month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar") %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point() + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw()

@@

temp_2002_2022 %>% filter(month_name == "Apr" | month_name == "May" | month_name == "Jun" | month_name == "Jul") %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point() + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw()

@@

temp_2002_2022 %>% filter(month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov") %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point() + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw()

@@

pollution %>% ggplot(
  aes(x = year, y = mean_pollution, color = pollutant)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution, color = pollutant)) + geom_point() + 
  labs(color="Pollutant") + xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = year, y = mean_pollution, factor(month_name, level = level_order))) +
  geom_point() + labs(color="Month") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2" ) %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

Carbon monoxide is a product of incomplete combustion of hydrocarbon-based fuels. In figure 12 and 13 we see a similar pattern as NO and NO2 where the level of pollution is decreasing throughout the years from 2002 to 2020 and the highest levels are in the winter months. The highest level of pollution for CO was seen in 2002 with 1762.5 ppb. This level is much higher than any of the other pollutants. This suggests that carbon monoxide levels are much more widespread than any of the other pollutants we have analyzed and due to this can have a affect on our health. Even with the decreasing levels, the affects will still be dangerous. The highest level in 2020 was around 550 ppb which is almost 5 times higher than the levels of NO, NO2, and O3 in 2020.

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

#### Visualizing Mean Pollution by Temperature and Pollutant

@@

pollu_temp %>% filter((pollutant != "CO" & pollutant != "SO2")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point() + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()
```

@@

pollu_temp %>% filter((pollutant == "CO")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point() + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()
```

@@

pollu_temp %>% filter((pollutant == "SO2")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point() + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()

@@

Carbon monoxide can significantly impact human health by entering the blood stream through the lungs and therefore inhibiting the blood's capacity to carry oxygen to organs and tissues. CO is harmful to those with heart disease and those with respiratory conditions. It also affects healthy individuals by reducing their exercise capacity, visual perception, manual dexterity, learning functions, and ability to perform complex tasks.
